A thermodynamic view of dusty protoplanetary disks

### 2017-09-23 summary.tex
update to re-submitted version, unbolded

{\bf

@@

\subsection{Implications for numerical simulations}\label{dust_sims}
{\bf The one-fluid model with the terminal velocity approximation, Eq. \ref{masseq}---\ref{momeq}, 
has been applied to simulate dusty protoplanetary disks \citep{dipierro15,ragusa17}. These studies 
explicitly evolve the dust density.  
}

{\bf However, the dusty/adiabatic gas equivalence identified in this paper
means that one need not implement a separate module for simulating the dust component in
locally isothermal/polytropic gas. The standard energy equation substitutes for the dust continuity 
equation. Thus pure gas dynamic codes can be used to simulate protoplanetary dusty disks. 
} %means that pure 
%gas dynamics codes can be applied to simulate locally 
%isothermal/polytropic dusty gas. One need not implement a separate 
%equation for the dust component: 
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
% the standard energy equation substitutes for the dust continuity
%equation. 
When {\bf the sound-speed} $c_s$ (or $K$) is not constant and/or the dust-gas coupling is imperfect, $\tstop\neq0$, one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal 
conduction, which is also common in modern hydrodynamics codes.

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. {\bf In fact,
  our code is unaware of the fact that it is modeling dusty gas.} 
We will apply 
this modified code ---d\textsc{Pluto}--- to study dusty disk-planet interaction (Lin et al., 
in preparation). 
Preliminary results {\bf show our approach} %indicate the simplified one-fluid 
reproduce features such as dusty rings associated with 
planet-induced gaps similar to that obtained from explicit two-fluid
simulations \citep[e.g.][]{dong17}.

@@

\section{Summary}\label{summary}
{\bf
In this paper, we examine the conditions under which the presence of
dust can trigger instabilities in gaseous protoplanetary disks.  To this end, 
we develop an analogy between isothermal  dusty-gas and pure ideal
gas. 
} 
The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, with stopping time $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy conservation 
following an adiabatic, pure gas.

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also show that radial variations in
  $\rhod/\rhog$ can trigger a type of VSI, even when the 
  usual sources of vertical shear -- vertical dust gradients and radial 
  temperature gradients -- are negligible.  
  %effectively translates to a non-uniform radial temperature
  %profile, and the resulting vertical shear can lead to instability,
  %even if the true sound-speed is constant. 
}

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_\mathrm{g}$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_\mathrm{g}$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop, ][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}

>>>>

{

@@

\subsection{Implications for numerical simulations}\label{dust_sims}
{ The one-fluid model with the terminal velocity approximation, Eq. \ref{masseq}---\ref{momeq}, 
has been applied to simulate dusty protoplanetary disks \citep{dipierro15,ragusa17}. These studies 
explicitly evolve the dust density.  
}

{ However, the dusty/adiabatic gas equivalence identified in this paper
means that one need not implement a separate module for simulating the dust component in
locally isothermal/polytropic gas. The standard energy equation substitutes for the dust continuity 
equation. Thus pure gas dynamic codes can be used to simulate protoplanetary dusty disks. 
} %means that pure 
%gas dynamics codes can be applied to simulate locally 
%isothermal/polytropic dusty gas. One need not implement a separate 
%equation for the dust component: 
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
% the standard energy equation substitutes for the dust continuity
%equation. 
When { the sound-speed} $c_s$ (or $K$) is not constant and/or the dust-gas coupling is imperfect, $\tstop\neq0$, one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal 
conduction, which is also common in modern hydrodynamics codes.

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. { In fact,
  our code is unaware of the fact that it is modeling dusty gas.} 
We will apply 
this modified code ---d\textsc{Pluto}--- to study dusty disk-planet interaction (Lin et al., 
in preparation). 
Preliminary results { show our approach} %indicate the simplified one-fluid 
reproduce features such as dusty rings associated with 
planet-induced gaps similar to that obtained from explicit two-fluid
simulations \citep[e.g.][]{dong17}.

@@

\section{Summary}\label{summary}
{
In this paper, we examine the conditions under which the presence of
dust can trigger instabilities in gaseous protoplanetary disks.  To this end, 
we develop an analogy between isothermal  dusty-gas and pure ideal
gas. 
} 
The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, with stopping time $\tstop \to 0$,  
 dust is entrained in 
the gas. Then the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the flow. This property is analogous to entropy conservation 
following an adiabatic, pure gas.

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. { We also show that radial variations in
  $\rhod/\rhog$ can trigger a type of VSI, even when the 
  usual sources of vertical shear -- vertical dust gradients and radial 
  temperature gradients -- are negligible.  
  %effectively translates to a non-uniform radial temperature
  %profile, and the resulting vertical shear can lead to instability,
  %even if the true sound-speed is constant. 
}

%implications for realistic disks
{ In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_\mathrm{g}$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_\mathrm{g}$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop, ][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}
